---
title: "Introduction to frailtoys"
author: "Theodor Balan"
date: "1/31/2017"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## 

## The theory {.smaller}
For cluster $i$, individual $j$, observation $k$ we have a corresponding intensity
$$
\lambda_{ij} = Y_{ij}(t) z_i\exp(\gamma'\mathbf{x}_{ij}(t)) \lambda_0(t)
$$
Each individual is a collection of observations of type `(tstart, tstop, status)`. From each of these lines, we have the cumulative intensity
$$
\Lambda_{ijk} = z_i\exp(\gamma'\mathbf{x}_{ijk}) \Lambda_{0, ijk}
$$
and for a cluster, we have the cumulative intensity in this cluster
$$
\Lambda_i=z_i \sum_{j,k}\exp(\gamma'\mathbf{x}_{ijk}) \Lambda_{0, ijk}
$$

## The theory
The conditional likelihood (given $z_i$) is then
$$
l_i = \prod_i \prod_{j,k} \left(\lambda_0(t_k) \exp(\gamma'\mathbf{x}_{ijk})\right)^{\delta_{ijk}} z_i^{n_i} \exp(-z_i\Lambda_i)
$$
The marginal likelihood is
$$
\tilde L_i = \prod_i \prod_{j,k} \left(\lambda_0(t_k) \exp(\gamma'\mathbf{x}_{ijk})\right)^{\delta_{ijk}} \mathrm{E}\left[ z_i^{n_i} \exp(-z_i\Lambda_i)  \right]
$$
where the expectation is 
$$
\mathrm{E}\left[ z_i^{n_i} \exp(-z_i\Lambda_i) \right] \equiv(-1)^{n_i} \mathcal{L}^{(n_i)}(\Lambda_i) 
$$

## The theory
Question is: what distribution should we take the expectation with regards to? 

We consider infinitely divisible random variables, with the Laplace transform
$$
\mathcal{L}(c; \alpha, \beta) = \exp\left(-\alpha \psi(c;\beta)\right)
$$

This includes the following:

  - gamma $\psi(c;\beta) = \log(c + \beta) - \log(\beta)$
  - positive stable $\psi(c; \beta) = \exp(c^\beta)$
  - PVF (Hougaard) with $\psi(c; \beta)$
  - inverse Gaussian (particular case of PVF with $m = -0.5$)

## Ascertainment
Consider that we have a cluster with only one observation / individual. Therefore, if the cluster has $n$ individuals, consider the ascertainment event
$$
A_i = A_{i1} \cap A_{i2} \cap...\cap A_{in}
$$
The distribution of the frailty in this cluster is then given by

$$
f_{Z|A_i}(z) = \frac{P(A_i|z) f(z)}{\int_z P(A_i|z) f(z) dz} = \frac{P(A_i|z)f(z)}{E_Z[P(A_i|z)]}
$$
Now this has Laplace transform
$$
\mathcal{L}_{Z|A_i}(c) = \frac{E_Z[e^{-cz}P(A_i|z)]}{E_Z[P(A_i|z)]}
$$

## Left truncation
With left truncation, with each individual truncated at $t_{L, ij}$ we have
$$
P(A_{ij}|z_i) = P(T_{ij}>t_{L,ij}|z_i) = \exp(-z_i \Lambda_{L,ij})]
$$
We then denote 
$$
\Lambda_{L,i} = \sum_j \Lambda_{L,ij}
$$
which would be the cumulative hazard from 0 to the entry time $t_{L,ij}$. 

## Left truncation
In that case we see that
$$
\mathcal{L}_{Z|A_i}(c) = \frac{E_Z[e^{-z (c + \Lambda_{L,i})}]}{E_Z[e^{-z\Lambda_{L,i}}]} = \frac{\mathcal{L}(c + \Lambda_{L,i})}{\mathcal{L}(\Lambda_{L,i})}
$$
In the infinite divisible framework we have
$$
\mathcal{L}_{Z|A_i}(c) = \frac{\exp(-\alpha \psi(c + \Lambda_{L,i}))}{\exp(-\alpha \psi(\Lambda_{L,i}))} = \exp(-\alpha \tilde \psi(c))
$$
with $\tilde \psi(c) = \psi(c + \Lambda_{L,i}) - \psi(\Lambda_{L,i})$.

## Left truncation
In the proportional intensity model, the hazard $\Lambda_L$ is not of interest, because it cancels out in the likelihood.

Here, it is of interest! 

Intuition: the frailty of an individual who survived until a certain time is not the same as the frailty of an individual at time 0. Survivors are more likely to be low-frailty individuals.

## More complicated ascertainment
In Mar's case, the ascertainment scheme was more complicated: at least two individuals must have survived until some time. Then $P(A_i)$ is not a simple function of the $P(A_{ij})$. 

## What Laplace transform?
It turns out that, under left truncation:

  - the frailty of survivors is still gamma with a change of parameters
  - stable and inverse gaussian become some distribution within the infinite divisible family
  - the PVF (Hougaard) distribution becomes some distribution within the infinite divisible family
  
## Problem with this
An observation line is usually `(tstart, tstop, status)` with an `id` specification for the cluster.

  - If we have recurrent events, then `tstart` refers to the time of a previous event, not left truncation
  - If we have time-dependent covariates, then `tstart` refers to a change in covariate values, not left truncation
  - If we have clustered data, then `tstart` *does* refer to left truncation

There is no way of distinguishing this automatically. 

## Welcome to frailtoys!
```{r}
library(frailtoys)
```

```{r}
# demo page





## Jeebus!



```
